### 2022-03-18 omicron-generation.tex
update output

Estimating the differences in the incubation-period and serial-interval distributions of SARS-CoV-2 variants is critical to understanding changes in their transmission dynamics and therefore controlling their outbreaks.
However, these comparisons are inherently difficult due to their dependence on epidemic growth rates---for example, when an epidemic is growing exponentially, a cohort of infected individuals who developed symptoms at the same time are more likely to have shorter incubation periods.
Here, we re-analyze incubation-period and serial-interval data collected by Backer \textit{et al.}, describing transmissions of the Delta and Omicron variants from Netherlands at the end of December 2021.
Previous analysis of the same data set reported shorter mean observed incubation period (3.2 days vs 4.4 days) and serial interval (3.5 days vs 4.1 days) for the Omicron variant, but the number of infections caused by the Delta variant decreased during this period as the number of Omicron infections increased.
When we account for growth-rate differences of two variants, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.3 days; 95\% CI: 2.8--3.7 days) than for the Delta variant (4.2 days; 95\% CI: 4.0--4.4 days).
The differences in generation intervals may reflect higher reproduction numbers of the Omicron variant, which can drive faster susceptible depletion among close contacts.
Neglecting differences in the generation-interval distributions can further bias in estimates of transmission advantages of the Omicron variant.

@@

Our study also has important implications for estimating transmission advantages of new SARS-CoV-2 variants.
In the example we consider, neglecting differences in the generation-interval distributions leads to $\approx 10\%$ bias in the estimates of the reproduction number ratios; however, the degree of ratio is expected to be sensitive to the assumed generation-interval distribution of the resident variant.
For example, \cite{pearson2021bounding} estimated a much higher reproduction ratios ($> 4$ fold) in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
With our generation-interval estimates, we estimate that the reproduction number ratio of 2.8 assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction number ratios with the estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
We also show that both growth-rate differences and the reproduction number ratios decreased over time---this result further corroborates are earlier theoretical framework, which showed that growth-rate differences changes over the course of an epidemic even when the reproduction number ratios remain constant \citep{park2021roles}.

>>>>

Estimating the differences in the incubation-period and serial-interval distributions of SARS-CoV-2 variants is critical to understanding changes in their transmission dynamics and therefore controlling their outbreaks.
However, these comparisons are inherently difficult due to their dependence on epidemic growth rates---for example, when an epidemic is growing exponentially, a cohort of infected individuals who developed symptoms at the same time are more likely to have shorter incubation periods.
Here, we re-analyze incubation-period and serial-interval data collected by Backer \textit{et al.}, describing transmissions of the Delta and Omicron variants from Netherlands at the end of December 2021.
Previous analysis of the same data set reported shorter mean observed incubation period (3.2 days vs 4.4 days) and serial interval (3.5 days vs 4.1 days) for the Omicron variant, but the number of infections caused by the Delta variant decreased during this period as the number of Omicron infections increased.
When we account for growth-rate differences of two variants, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.0 days; 95\% CI: 2.7--3.2 days) than for the Delta variant (3.8 days; 95\% CI: 3.7--4.0 days).
The differences in generation intervals may reflect higher reproduction numbers of the Omicron variant, which can drive faster susceptible depletion among close contacts.
Neglecting differences in the generation-interval distributions can further bias in estimates of the reproduction advantage of the Omicron variant.

@@

Our study also has important implications for estimating transmission advantages of new SARS-CoV-2 variants.
In the example we consider, neglecting differences in the generation-interval distributions leads to $\approx 10\%$ bias in the estimates of the reproduction number ratios; however, the degree of ratio is expected to be sensitive to the assumed generation-interval distribution of the resident variant.
For example, \cite{pearson2021bounding} estimated a much higher reproduction ratios ($> 4$ fold) in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
With our generation-interval estimates, we estimate that the reproduction number ratio of 2.6 assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction number ratios with the estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
We also show that both growth-rate differences and the reproduction number ratios decreased over time---this result further corroborates are earlier theoretical framework, which showed that growth-rate differences changes over the course of an epidemic even when the reproduction number ratios remain constant \citep{park2021roles}.
